---
title: "Rmarkdown_Moquete#3a"
author: "Renz Moquete"
date: "2025-10-12"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. There is a built-in vector LETTERS contains the uppercase letters of the alphabet
### and letters which contains the lowercase letters of the alphabet.


```{r}
alphabet <- LETTERS
alphabet

small_alphabet <- letters
small_alphabet
```

### a. You need to produce a vector that contains the first 11 letters.

```{r}
alphabet <- LETTERS
alphabet[1:11]

```



### b. Produce a vector that contains the odd numbered letters.

```{r}
alphabet <- LETTERS
odd <- alphabet[seq(1, length(alphabet), 2)]
odd
```

### c. Produce a vector that contains the vowels

```{r}
alphabet <- LETTERS
vowels <- alphabet [c(1, 5, 9, 15, 21)]
vowels

```

### d. Produce a vector that contains the last 5 lowercase letters.

```{r}
alphabet <- letters
last_fivelower <- tail(alphabet, 5)
last_fivelower

```


### e. Produce a vector that contains letters between 15 to 24 letters in lowercase.

```{r}
small_alphabet <- letters
small_alphabet[15:24]

```


### 2. Create a vector(not a dataframe) with the average temperatures in April for Tuguegarao City, Manila, Iloilo City, Tacloban, Samal Island, ### and Davao City. The average temperatures in Celcius are 42, 39, 34, 34, 30, and 27 degrees.

```{r}
cities <- c("Tuguegarao City", "Manila", "Iloilo City", "Tacloban", "Samal Island", "Davao City")
temp <- c(42, 39, 34, 34, 30, 27)
cities
temp
```

### 3. a. What is the R code and its result for creating a character vector for the city/town
## of Tuguegarao City, Manila, Iloilo City, Tacloban,Samal Island, and Davao City? Name the
## object as city. The names should follow the same order as in the instruction.

``` {r}
cities <- c("Tuguegarao City", "Manila", "Iloilo City", "Tacloban", "Samal Island", "Davao City")

```

### b. The average temperatures in Celcius are 42, 39, 34, 34, 30, and 27 degrees.
### Name the object as temp. Write the R code and its output. Numbers should also follow
### what is in the instruction.

``` {r}
temp <- c(42, 39, 34, 34, 30, 27)

```

### c. Create a dataframe to combine the city and the temp by using ‘data.frame(). What
### the R code and its result?

``` {r}
cities <- c("Tuguegarao City", "Manila", "Iloilo City", "Tacloban", "Samal Island", "Davao City")
temp <- c(42, 39, 34, 34, 30, 27)

frame = data.frame(cities, temp)
frame
```
### d. Associate the dataframe you have created in 2.(c) by naming the columns using
### the names() function. Change the column names by using names() function as City and
### Temperature. What is the R code and its result?


``` {r}
cities <- c("Tuguegarao City", "Manila", "Iloilo City", "Tacloban", "Samal Island", "Davao City")
temp <- c(42, 39, 34, 34, 30, 27)

frame = data.frame(cities, temp)
names(frame) <- c("City", "Temperature")
frame
```


### e. Print the structure by using str() function. Describe the output.
``` {r}
cities <- c("Tuguegarao City", "Manila", "Iloilo City", "Tacloban", "Samal Island", "Davao City")
temp <- c(42, 39, 34, 34, 30, 27)

frame = data.frame(cities, temp)
names(frame) <- c("City", "Temperature")
frame

str(frame) 

```

### f. From the answer in d, what is the content of row 3 and row 4 What is its R code and
## its output?

``` {r}
cities <- c("Tuguegarao City", "Manila", "Iloilo City", "Tacloban", "Samal Island", "Davao City")
temp <- c(42, 39, 34, 34, 30, 27)

frame = data.frame(cities, temp)
names(frame) <- c("City", "Temperature")
frame[3:4, ]


```

### g. From the answer in d, display the city with highest temperature and the city with
### the lowest temperature. What is its R code and its output?

``` {r}
maximum <- max(frame$Temperature)
minimum <- min(frame$Temperature)

frame[frame$Temperature == maximum, ]
frame[frame$Temperature == minimum, ]
```

### Using Matrice

### 2. Create a matrix of one to eight and eleven to fourteen with four columns and three rows
#### a. What will be the R code for the #2 question and its result?

``` {r}
matrice <- matrix(c(1,2,3,4,5,6,7,8,11,12,13,14),nrow = 3, byrow = TRUE)
matrice
```

### b. Multiply the matrix by two. What is its R code and its result?
``` {r}
matrice <- matrix(c(1,2,3,4,5,6,7,8,11,12,13,14),nrow = 3, byrow = TRUE)
double_matrice <- matrice * 2
double_matrice
```

### c. What is the content of row 2? What is its R code?

``` {r}
matrice[2, ]
```


### d. What will be the R code if you want to display the column 3 and column 4 in row 1 and row 2? What is its output?

``` {r}
matrice[1:2, 3:4]
```


### e. What is the R code is you want to display only the columns in 2 and 3, row 3? What is its output?

``` {r}
matrice[3, 2:3]
```

### f. What is the R code is you want to display only the columns 4? What is its output?
``` {r}
matrice[ , 4]
```

### g. Name the rows as isa, dalawa, tatlo and columns as uno, dos, tres, quatro
### for the matrix that was created in b.‘. What is its R code and corresponding output?

``` {r}
matrice <- matrix(c(1,2,3,4,5,6,7,8,11,12,13,14),nrow = 3, byrow = TRUE)
rownames(matrice) <- c("isa", "dalawa", "tatlo")
colnames(matrice) <- c("uno", "dos", "tres", "quatro")
matrice
```

### h. From the original matrix you have created in a, reshape the matrix by assigning a
### new dimension with dim(). New dimensions should have 2 columns and 6 rows. What will
### be the R code and its output?

``` {r}
dim(matrice) <- c(6, 2)
matrice
```


### 3. An array contains 1, 2, 3, 6, 7, 8, 9, 0, 3, 4, 5, 1

``` {r}
array_dta <- c(1, 2, 3, 6, 7, 8, 9, 0, 3, 4, 5, 1)
array_dta
```

### a. Create an array for the above numeric values. Each values will be repeated twice
### What will be the R code if you are to create a three-dimensional array with 4 columns and 2 rows. What will be its output?


```{r} 
num = array(array_dta, dim = c(2,4,3))
num
```

### b. How many dimensions do your array have?
### There are total of 3 dimensions.

### c. Name the rows as lowercase letters and columns as 
### uppercase letters starting from the A. The array names should be 
### “1st-Dimensional Array”,  “2nd-Dimensional Array”, and
### “3rd-Dimensional Array”. What will be the R codes and its ##output

```{r} 
dimnames(num) <- list( letters[1:2],  LETTERS[1:4], c("1st-Dimensional Array", "2nd-Dimensional Array", "3rd-Dimensional Array")
)
num

```


