---
title: "RWorksheet_Moquete#4a.Rmd"
author: "Renz Moquete"
date: "2025-11-11"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### 1. The table below shows the data about shoe size and height. Create a data frame.

```{r}
shoe_size <- c(6.5, 9.0, 8.5, 8.5, 10.5, 7.0, 9.5, 9.0, 13.0, 7.5, 10.5, 8.5, 12.0, 10.5, 13.0, 11.5, 8.5, 5.0, 10.0, 6.5, 7.5, 8.5, 10.5, 8.5, 10.5, 11.0, 9.0, 13.0)
height <- c(66.0, 68.0, 64.5, 65.0, 70.0, 64.0, 70.0, 71.0, 72.0, 64.0, 74.5, 67.0, 71.0, 71.0, 77.0, 72.0, 59.0, 62.0, 72.0, 66.0, 64.0, 67.0, 73.0, 69.0, 72.0, 70.0, 69.0, 70.0)
gender <- c('F', 'F', 'F', 'F', 'M', 'F', 'F', 'F', 'M', 'F', 'M', 'F', 'M', 'M', 'M', 'M', 'F', 'F', 'M', 'F', 'F', 'M', 'M', 'F', 'M', 'M', 'M', 'M')

# Combine the vectors into a data frame
shoe_height_data <- data.frame(
  Shoe_Size = shoe_size,
  Height = height,
  Gender = gender
)
```

### a. Describe the data
The data shows the an organize table of data such as shoe size, height and gender.

### b. Create a subset by males and females with their corresponding shoe size and height.
What its result? Show the R scripts.

```{r}
males_data <- subset(shoe_height_data, Gender == 'M')

# Create a subset for Females
females_data <- subset(shoe_height_data, Gender == 'F')

# Print the results to the console
print("Male Subset:")
print(males_data)

print("Female Subset:")
print(females_data)
```
### c. Find the mean of shoe size and height of the respondents. Write the R scripts and its result.

```{r}
mean_shoe_size <- mean(shoe_height_data$Shoe_Size)

# Calculate the mean of the 'Height' column
mean_height <- mean(shoe_height_data$Height)

# Print the results to the console
mean_height
mean_shoe_size
```

### d. Is there a relationship between shoe size and height? Why?

Answer: Yes, there is a strong relationship between shoe size and height.

```{r}
mean_shoe_size <- mean(shoe_height_data$Shoe_Size)

# Calculate the mean of the 'Height' column
mean_height <- mean(shoe_height_data$Height)

# Print the results to the console
mean_height
mean_shoe_size
```

### 2. Construct character vector months to a factor with factor() and assign the result to
factor_months_vector. Print out factor_months_vector and assert that R prints out
the factor levels below the actual values.

```{r}
# 1. Create a character vector with the names of the months
months_vector <- c("March","April","January","November","January",
                   "September","October","September","November","August",
                   "January","November","November","February","May","August",
                   "July","December","August","August","September","November",
                   "February","April")

# 2. Convert the character vector to a factor
factor_months_vector <- factor(months_vector)

# 3. Print the resulting factor to the console
print(factor_months_vector)
```

### 3. Then check the summary() of the months_vector and factor_months_vector. | Interpret the results of both vectors. Are they both equally useful in this case?

```{r}
summary(months_vector)
summary(factor_months_vector)
```

Interpretation: The summary of a character vector is purely structural. It tells you what the data is (text) and how much of it there is (24 items), but it provides zero insight into the content of the data. It doesn't know that "January" appeared multiple times or how many unique months there are. For data analysis, this summary is not very useful.